From Trends to Truth: A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates

### 2024-02-14 paper/paper.qmd
minor tweaks to paper

@fig-1 is the replication of Figure 1 from `The Puzzle of Falling US Birth Rates since the Great Recession' by @citeOriginalReplication, and showcases the trajectory of US birth rates spanning from 1980 to 2020, focusing on births per 1000 women aged 15 to 44. The peak occurred in 1990, reaching 70.9 births per 1000 women, closely followed by 2007 with 69.3 births per 1000 women. In contrast, 2020 marked the lowest point with only 55.8 births per 1000 women. From 1980 to 2007, the birth rate experienced moderate fluctuations, with a maximum variance of 7.3 births per 1000 women. However, following the year 2007, the birth rate seems to display a consistent a downward trend, declining by 13.5 births per 1000 women between 2007 and 2020.

@@

@fig-2a is the replication of Figure 2 (a) from the original paper. Examining the trends in birth rates by five-year age groups reveals intriguing patterns.

In the 15 - 19 years old age group, there has been a consistent decline in birth rates, punctuated by a brief period of growth from 1986 to 1991, peaking at 61.8 births per 1000 women in 1991 before steadily declining to its lowest recorded rate of 15.3 births per 1000 women in 2020.

The 20-24 years old age group remained relatively stable from 1980 to 2007, reaching a peak of 116.5 births per 1000 women in 1990. However, it has experienced a consistent decline from 2007 onward, reaching a low of 62.8 births per 1000 women in 2020.

Similarly, the 25-29 years old age group displayed stability until 2007, with the highest birth rate of 120.2 births per 1000 women recorded in 1990. Despite a gradual decline post-2007, it retained a relatively higher birth rate compared to younger age groups, hitting a low of 90.0 births per 1000 women in 2020.

@@

The 40-44 years old age group displayed minimal change, remaining relatively steady over the years, with slight increases noted annually. The lowest birth rate was recorded in 1980 at 8.1 births per 1000 women, rising to 12.0 births per 1000 women in 2019.

Analyzing the data in @fig-2a, the 25 - 29 years old age group consistently led in birth rates until being surpassed by the 30 - 34 years old age group in 2016. Conversely, the 40-44 years old age group consistently exhibited the lowest birth rates. However, with declining births in the 15-19 years old age group and a stable, increasing trend in the 40-44 years old age group, a shift in the age group with the lowest birth rate may soon occur for the first time in 40 years.

Additionally, a noteworthy trend is the steady decline in birth rates among younger age groups from around 2007 onwards, while the older age groups have experienced the opposite trend.

>>>>

@fig-1 is the replication of Figure 1 from `The Puzzle of Falling US Birth Rates since the Great Recession' by @citeOriginalReplication, and showcases the trajectory of US birth rates spanning from 1980 to 2020, focusing on births per 1000 women aged 15 to 44. The peak occurred in 1990, reaching 70.9 births per 1000 women, closely followed by 2007 with 69.3 births per 1000 women. In contrast, 2020 marked the lowest point with only 55.8 births per 1000 women which is nearly a 21.3% decrease in births since 1990. From 1980 to 2007, the birth rate experienced moderate fluctuations, with a maximum variance of 7.3 births per 1000 women. However, following the year 2007, the birth rate seems to display a consistent a downward trend, declining by 13.5 births per 1000 women or a 19.5% decrease between 2007 and 2020.

@@

Next, @fig-2a is the replication of Figure 2 (a) from the original paper. Concerning itself with examining the trends in birth rates by five-year age groups, we can observe insightful results.

Looking at the 15 - 19 years old age group, there has been a consistent decline in birth rates, punctuated by a brief period of growth from 1986 to 1991, peaking at 61.8 births per 1000 women in 1991 before steadily declining to its lowest recorded rate of 15.3 births per 1000 women in 2020.

The 20-24 years old age group remained relatively stable from 1980 to 2007, reaching a peak of 116.5 births per 1000 women in 1990. However, similarly to the previous age group, it experiences a consistent decline from 2007 onward, reaching a low of 62.8 births per 1000 women in 2020.

Again, the 25-29 years old age group displays stability until 2007, with the highest birth rate of 120.2 births per 1000 women recorded in 1990. Despite a gradual decline post-2007, it retained a relatively higher birth rate compared to younger age groups, hitting a low of 90.0 births per 1000 women in 2020.

@@

Lastly, the 40-44 years old age group displayed minimal change, remaining relatively steady over the years, with slight increases noted annually. The lowest birth rate was recorded in 1980 at 8.1 births per 1000 women, rising to 12.0 births per 1000 women in 2019.

Observing the data in @fig-2a, we find the 25 - 29 years old age group consistently led in birth rates until being surpassed by the 30 - 34 years old age group in 2016. Conversely, the 40-44 years old age group consistently exhibited the lowest birth rates. However, with declining births in the 15-19 years old age group and a stable, increasing trend in the 40-44 years old age group, a shift in the age group with the lowest birth rate may soon occur for the first time in 40 years.

Additionally, a noteworthy trend is the steady decline in birth rates among younger age groups from around 2007 onward, while the older age groups have experienced the opposite trend.
